---
title: "Masters Homework 2"
author: "Alessia Toska"
date: "7/21/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo = TRUE}
nyc <- read.csv("https://raw.githubusercontent.com/fredlapolla/RVilcekMaster23/master/NYC_HANES_DIAB.csv")
```
>- Find summary information about the total blood mercury levels in our NYC HANES data (hint: names(nyc) will call the names of each column). What is the max value, what is the minimum value.  

```{r}
summary(nyc$MERCURYTOTALBLOOD)
max(nyc$MERCURYTOTALBLOOD, na.rm=T)
min(nyc$MERCURYTOTALBLOOD, na.rm=T)
```

>- Bonus/Challenge: What is the Standard Deviation and skewness of the blood mercury?

```{r}
#standard deviation
sd(nyc$MERCURYTOTALBLOOD, na.rm = TRUE)
```
#skewness manually
```{r}
x <- na.omit (nyc$MERCURYTOTALBLOOD)

n <- length(x)
mean_x <- mean(x)
sd_x <- sd(x)

skewness_formula<- (n * sum((x - mean_x)^3)) / ((n - 1) * (n - 2) * sd_x^3)

skewness_formula
```


>- Install a package and load, such as "ggplot2" or "dplyr"

```{r}
install.packages("ggplot2")
library(ggplot2)
```

>- What two commands are needed to install and use a package in R?

#example with dplyr
```{r}
install.packages("dplyr")
library(dplyr)
```



>- When you read in a .csv file into R, what class is the object that you read in (like nyc)?

A data frame


>- What command can you use to determine what type of data is in a column?

```{r}
class(nyc$GENDER) #taking gender as an example
```


>- Provide a command to find the type of data in the DX_DBTS column of nyc. 

```{r}
class(nyc$DX_DBTS)
```


>- Provide the same command the type of data but use the other way of indexing from what you did above. 

```{r}
class(nyc[, "DX_DBTS"])
```


>- What would happen if we had chosen the first 4 columns of NYC HANES and made a matrix? What would the class of the cells in this matrix be? Why?

```{r}
mini_matrix <- as.matrix(nyc[, 1:4])
class(mini_matrix)
```
R shows both classes because a matrix is simply a 2 dimensional array in R’s class hierarchy, so it inherits the general array class alongside the more specific matrix class.

>- If we make a vector,vec<- c(3,9,4,7,8,9), and multiply it by WannaMatrix, what is the value in the third row, third column? 

```{r}
WannaMatrix <- nyc[5:10, 6:9]
WannaMatrix <- as.matrix(WannaMatrix,nrow = 6, ncol = 4)

vec <- c(3, 9, 4, 7, 8, 9)

new_mat <- WannaMatrix * vec 

new_mat[3, 3]  

```



>- What happens if instead we made a vector vec2 <- c(3,9,4,7,8) and multiply it by WannaMatrix, what is the third row, third column value? What is going on? 

```{r}
vec2 <- c(3,9,4,7,8)
new_mat2 <- WannaMatrix * vec2

new_mat2[3, 3]  
```


>- Look up the arguments for read.csv

```{r}
?read.csv

#or

args(read.csv)
```


>- Use apply() to find the median value of CADMIUM, LEAD CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, and CREATININE

```{r}
cols<- c("CADMIUM", "LEAD", "CHOLESTEROLTOTAL", "GLUCOSESI", "CREATININESI", "CREATININE")

median_all <- apply(nyc[, cols], 2, median, na.rm = TRUE)

median_all
```

>- Do the same using sapply()

```{r}
sapply_median <- sapply(nyc[, cols], median, na.rm = TRUE)
sapply_median
```

>- Use a for loop to find the median value of columns 16:19

```{r}
for( i in 16:19) {
  value <- median(nyc[,i], na.rm = TRUE)
  print(c(colnames(nyc[i]), value))
}
```

>- Make a for loop to find the standard deviation (hint sd() ) of CHOLESTEROLTOTAL, GLUCOSESI, CREATININESI, AND CREATININE

```{r}
for (i in 16:19) {
  sd_value <-sd(nyc[,i], na.rm = TRUE)
  print(c(colnames(nyc)[i], sd_value))
}
```


>- Bonus/Challenge Create a vector in which to put the standard deviations 

```{r}
st_dev <- sapply(nyc[16:19], sd, na.rm = TRUE)
st_dev

#is.vector(st_dev) #just checking
```


>- Use ifelse to create a new variable of high and normal A1C dichotomizing with 7 as a threshold for high levels

```{r}
nyc$A1C_status <- ifelse(nyc$A1C > 7, "high", "normal")
#head(nyc$A1C_status) #just checking
```


>- Subset to only have females (gender == "female")

```{r}
nyc$GENDER <-factor(nyc$GENDER, labels = c("male", "female"))
nyc_female <- subset(nyc, GENDER == "female") #im assuming the females are gender 2 
```


>- Create a subset of males and find the median total cholesterol level

```{r}
nyc_male <- subset(nyc, GENDER == "male")
median(nyc_male$CHOLESTEROLTOTAL, na.rm = TRUE)
```


>- Try to create a summary of the mean A1C using two factors, GENDER and DX_DBTS

```{r}
by(nyc$A1C,
   INDICES = list(Gender   = nyc$GENDER,
                  Diabetes = nyc$DX_DBTS),
   FUN     = mean,
   na.rm   = TRUE)
```

Find the class of the first three columns of data. What are they?

```{r}
sapply(nyc[ , 1:3], class)
```

